Estimating the genome-wide contribution of selection to temporal allele frequency change

### 2020-05-13 manuscript/manuscript.tex
editing seasonal section

We first analyzed \textcite{Barghi2019-qy}, an evolve-and-resequence study with
ten replicate populations exposed to a high temperature lab environment and
evolved for 60 generations, and sequenced every ten generations. Using the
seven timepoints and ten replicate populations, we estimated the genome-wide $6
\times 6$ temporal covariance matrix $\mathbf{Q}$ for each of the ten
replicates. Each row of these matrices represent the temporal covariance
$\cov(\Delta_{_{10}} p_s, \Delta_{_{10}} p_t)$, between the allele frequency
change (in ten-generation intervals, denoted $\Delta_{_{10}} p_t$) \vb{of} some
initial reference generation $s$ (the row of the matrix), and some later
timepoint $t$ (the column of the matrix). We corrected these matrices for
biases created due to sampling noise, and \vb{normalized} the entries for
heterozygosity (see Supplementary Materials Sections
\ref{supp:ind-depth-var-corr} and \ref{supp:matrix-correction}). These
covariances are expected to be zero when only drift is acting, as only
heritable variation for fitness can create covariance between allele frequency
changes in a closed population \parencite{Buffalo2019-io}.  Averaging across
the ten replicate temporal covariances matrices, we find temporal covariances
that are statistically significant (95\% block bootstraps CIs do not contain
zero), consistent with linked selection perturbing genome-wide allele frequency
changes over very short time periods. The covariances between all adjacent time
intervals are positive and then decay towards zero as we look at more distant
time intervals (Figure \ref{fig:figure-1} A), as expected when directional
selection affects linked variants' frequency trajectories until ultimately
linkage disequilibrium and the associated additive genetic variance for fitness
decays (\vb{which could occur as a population reaches a new optimum})
\parencite{Buffalo2019-io}. The temporal covariances per replicate are noisier
but this general pattern holds; see Supplementary Figure
\ref{suppfig:barghi-cov-panels}. \textcite{Barghi2019-qy}'s design means that
we cannot \vb{observe covariances between pairs of generations that constitute
the estimated covariances between ten-generation time intervals; had these
covariances been measurable on shorter timescales, their cumulative effect may
have been higher yet (see Supplementary Material Sections
\ref{supp:barghi-covs} and \ref{supp:tempblocks} for more details).}

Since our covariances are essentially averages over loci, the covariance
estimate could be strongly affected by a few outlier regions. To test whether
large outlier regions drive the genome-wide signal we see in the
\textcite{Barghi2019-qy} data, we calculate the covariances in 100kb windows
along the genome (we refer to these as windowed covariances throughout) and
take the median windowed covariance, and trimmed-mean windowed covariance, as a
measure of the genome-wide covariance robust to large-effect loci. These robust
estimates (Supplementary Table \ref{supp:table-trimmed-mean} and Supplementary
Figure \ref{suppfig:barghi-trimmed-mean}) confirm the patterns we see using the
mean covariance, confirming that genomic temporal covariances are non-zero due
to the impact of selection acting across many genomic \vb{regions}.

@@

which estimates the \vb{impact of selection} on allele frequency change between
the initial generation $0$ and some later generation $t$, which can be varied
to see how this quantity grows through time. When the sum of the covariances is
positive, this measure can intuitively be understood as the relative fraction
of allele frequency change normally thought of as ``drift" that is actually due
to selection. Additionally, $G(t)$ can be understood as a short-timescale
estimate of the reduction in neutral diversity due to linked selection (or
equivalently the reduction in neutral effective population size needed to
account linked selection, see Supplementary Materials Section
\ref{supp:reduction-approx}). Since \textcite{Barghi2019-qy} experiment is
sequenced every ten generations, \vb{the numerator uses the covariances
  estimated between ten-generation blocks of allele frequency change}; \vb{thus
  the strong, unobservable, covariances between adjacent generations do not
contribute to the numerator of $G(t)$}. Additionally, selection also inflates
the variance in allele frequency change per generation; however, this effect
cannot be easily distinguished from drift. For both these reasons, our measure
$G(t)$ is \vb{considerably} conservative (\vb{we demonstrate this through
simulations in Supplementary Material Section \ref{supp:tempblocks}}).  Still,
we find a remarkably strong signal. Greater than $20\%$ of total, genome-wide
allele frequency change over 60 generations is the result of selection (Figure
\ref{fig:figure-1} B). This proportion of variance attributable to selection
builds over time in Figure \ref{fig:figure-1}B as the effects of linked
selection are compounded over the generations unlike genetic drift. Our G(t)
starts to plateau to a constant level as the covariances from earlier
generations have decayed and so no longer contribute as strongly (Figure
\ref{fig:figure-1}).

>>>>

We first analyzed \textcite{Barghi2019-qy}, an evolve-and-resequence study with
ten replicate populations exposed to a high temperature lab environment and
evolved for 60 generations, and sequenced every ten generations. Using the
seven timepoints and ten replicate populations, we estimated the genome-wide $6
\times 6$ temporal covariance matrix $\mathbf{Q}$ for each of the ten
replicates. Each row of these matrices represent the temporal covariance
$\cov(\Delta_{_{10}} p_s, \Delta_{_{10}} p_t)$, between the allele frequency
change (in ten-generation intervals, denoted $\Delta_{_{10}} p_t$) \vb{of} some
initial reference generation $s$ (the row of the matrix), and some later
timepoint $t$ (the column of the matrix). We corrected these matrices for
biases created due to sampling noise, and \vb{normalized} the entries for
heterozygosity (see Supplementary Materials Sections
\ref{supp:ind-depth-var-corr} and \ref{supp:matrix-correction}). These
covariances are expected to be zero when only drift is acting, as only
heritable variation for fitness can create covariance between allele frequency
changes in a closed population \parencite{Buffalo2019-io}.  Averaging across
the ten replicate temporal covariances matrices, we find temporal covariances
that are statistically significant (95\% block bootstraps CIs do not contain
zero), consistent with linked selection perturbing genome-wide allele frequency
changes over very short time periods. The covariances between all adjacent time
intervals are positive and then decay towards zero as we look at more distant
time intervals (Figure \ref{fig:figure-1} A), as expected when directional
selection affects linked variants' frequency trajectories until ultimately
linkage disequilibrium and the associated additive genetic variance for fitness
decays (\vb{which could occur as a population reaches a new optimum})
\parencite{Buffalo2019-io}. The temporal covariances per replicate are noisier
but this general pattern holds; see Supplementary Figure
\ref{suppfig:barghi-cov-panels}.

Since our covariances are \gc{averages} over loci, the covariance
estimate could be strongly affected by a few outlier regions. To test whether
large outlier regions drive the genome-wide signal we see in the
\textcite{Barghi2019-qy} data, we calculate the covariances in 100kb windows
along the genome (we refer to these as windowed covariances throughout) and
take the median windowed covariance, and trimmed-mean windowed covariance, as a
measure of the genome-wide covariance robust to large-effect loci. These robust
estimates (Supplementary Table \ref{supp:table-trimmed-mean} and Supplementary
Figure \ref{suppfig:barghi-trimmed-mean}) confirm the patterns we see using the
mean covariance, confirming that genomic temporal covariances are non-zero due
to the impact of selection acting across many genomic \vb{regions}.

@@

%\textcite{Barghi2019-qy}'s design means that
%we cannot \vb{observe covariances between pairs of generations that constitute
%the estimated covariances between ten-generation time intervals
%
which estimates the \vb{impact of selection} on allele frequency change between
the initial generation $0$ and some later generation $t$, which can be varied
to see how this quantity grows through time. When the sum of the covariances is
positive, this measure can intuitively be understood as the relative fraction
of allele frequency change normally thought of as ``drift" that is actually due
to selection. Additionally, $G(t)$ can be understood as a short-timescale
estimate of the reduction in neutral diversity due to linked selection (or
equivalently the reduction in neutral effective population size needed to
account linked selection, see Supplementary Materials Section
\ref{supp:reduction-approx}). Since \textcite{Barghi2019-qy} experiment is
sequenced every ten generations, \vb{the numerator uses the covariances
  estimated between ten-generation blocks of allele frequency change}; \vb{thus
  the strong, unobservable, covariances between adjacent generations do not
  contribute to the numerator of $G(t)$}. \gc{Had these
covariances been measurable on shorter timescales, their cumulative
effect would likely have been higher yet (see Supplementary Material Sections
\ref{supp:barghi-covs} and \ref{supp:tempblocks} for more details).}
Additionally, selection also inflates
the variance in allele frequency change per generation; however, this effect
cannot be easily distinguished from drift. For both these reasons, our measure
$G(t)$ is \vb{considerably} conservative (\vb{we demonstrate this through
simulations in Supplementary Material Section \ref{supp:tempblocks}}).  Still,
we find a remarkably strong signal. Greater than $20\%$ of total, genome-wide
allele frequency change over 60 generations is the result of selection (Figure
\ref{fig:figure-1} B). This proportion of variance attributable to selection
builds over time in Figure \ref{fig:figure-1}B as the effects of linked
selection are compounded over the generations unlike genetic drift. Our G(t)
starts to plateau to a constant level as the covariances from earlier
generations have decayed and so no longer contribute as strongly (Figure
\ref{fig:figure-1}).
